DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life

### 2022-03-24 manuscript/figures-alt.tex
updating figures on ms

\newpage
\begin{figure}[!h]
\includegraphics{../figures/Fig1.pdf}
\caption{}
\label{fig:workflow}
\end{figure}
% \begin{center}
% \textsc{Figure \ref{fig:workflow}}
% \end{center}
Stylized DateLife workflow. This shows the general worflows and analyses that can be performed with \texttt{datelife}, via the R package or through the website  at \url{http://www.datelife.org/}. Details on the functions involved on each workflow are shown in \texttt{datelife}'s R package vignette.
\newpage
\begin{figure}[!h]
\includegraphics[width=1\linewidth]{../figures/fig_runtime_main.pdf}
\caption{}
\label{fig:runtime1}
\end{figure}
% \begin{center}
% \textsc{Figure \ref{fig:runtime1}}
% \end{center}
Computation time of query processing and search across \texttt{datelife}'s chronogram database relative to number of input taxon names. We sampled N names from the class Aves for each cohort 100 times and then performed a search with query processing not using the Taxon Names Resoultion Service (TNRS; dark gray), and using TNRS (light gray). We also performed a search using the already processed query for comparison (light blue).
\newpage
\begin{figure}[!h]
\includegraphics[width=1\linewidth]{../figures/fig_schronograms1.pdf}
\caption{}
\label{fig:schronograms1}
\end{figure}
% \begin{center}
% \textsc{Figure \ref{fig:schronograms1}}
% \end{center}
Lineage through time (LTT) plots of source chronograms containing all or a subset of species from the bird family Fringillidae of true finches. Arrows indicate maximum age of each chronogram. Numbers reference to chronograms' original publications 1: Barker et al. (\protect\hyperlink{ref-barker2012going}{2012}), 2: Barker et al. (\protect\hyperlink{ref-barker2015new}{2015}), 3: Burns et al. (\protect\hyperlink{ref-burns2014phylogenetics}{2014}), 4: Claramunt and Cracraft (\protect\hyperlink{ref-claramunt2015new}{2015}), 5: Gibb et al. (\protect\hyperlink{ref-gibb2015new}{2015}), 6: Hedges et al. (\protect\hyperlink{ref-Hedges2015}{2015}), 7: Hooper and Price (\protect\hyperlink{ref-hooper2017chromosomal}{2017}), 8: Jetz et al. (\protect\hyperlink{ref-Jetz2012}{2012}), 9: Price et al. (\protect\hyperlink{ref-price2014niche}{2014}).
\newpage
\begin{figure}[!h]
\includegraphics{../figures/fig_summaries.pdf}
\caption{}
\label{fig:summaries}
\end{figure}
% \begin{center}
% \textsc{Figure \ref{fig:summaries}}
% \end{center}
LTT plots of median (top) and Supermatrix Distance Method (SDM; bottom) chronograms summarising information from source chronograms found for the Fringillidae. Arrows indicate tree maximum age.
\newpage
\begin{figure}[!h]
\includegraphics{../figures/fig_crossval_bladj.pdf}
\caption{}
\label{fig:cvbladj}
\end{figure}
% \begin{center}
% \textsc{Figure \ref{fig:cvbladj}}
% \end{center}
LTT plots showing results from the cross-validation analyses of trees without branch lengths dated using BLADJ. The dating analysis can only be performed in trees with more than 2 tips, thus excluding chronogram from study 4; its data was still used as calibration for the other source chronograms.
\newpage
\begin{figure}[!h]
\includegraphics{../figures/fig_crossval_boldsumm.pdf}
\caption{}
\label{fig:cvbold}
\end{figure}
% \begin{center}
% \textsc{Figure \ref{fig:cvbold}}
% \end{center}
LTT plots showing results from the cross-validation analyses of trees with branch length reconstructed with data from the Barcode of Life Database (BOLD) dated using PATHd8. We could construct a tree with branch lengths for all source chronograms. However, dating with PATHd8 was only successful in three source chronograms shown here.

>>>>

\newpage
\begin{figure}[!h]
\includegraphics{../figures/figure1/figure1-vertical-final.pdf}
\caption{Stylized DateLife workflow. This shows the general worflows and analyses that can be performed with \texttt{datelife}, via the R package or through the website  at \url{http://www.datelife.org/}. Details on the functions involved on each workflow are shown in \texttt{datelife}'s R package vignette.
}
\label{fig:workflow}
\end{figure}
% \begin{center}
% \textsc{Figure \ref{fig:workflow}}
% \end{center}
\newpage
\begin{figure}[!h]
\includegraphics[width=1\linewidth]{../figures/fig_runtime_main.pdf}
\caption{Computation time of query processing and search across \texttt{datelife}'s chronogram database relative to number of input taxon names. We sampled N names from the class Aves for each cohort 100 times and then performed a search with query processing not using the Taxon Names Resoultion Service (TNRS; dark gray), and using TNRS (light gray). We also performed a search using the already processed query for comparison (light blue).}
\label{fig:runtime1}
\end{figure}
% \begin{center}
% \textsc{Figure \ref{fig:benchmark}}
% \end{center}
\newpage
\begin{figure}[!h]
\includegraphics[width=1\linewidth]{../figures/figure2/figure2-1.pdf}
\caption{Age data results of a DateLife search of a small sample of 6 bird species within the Passeriformes. Input names were found across 9 chronograms within 6 independent studies (Barker et al. (\protect\hyperlink{ref-barker2012going}{2012}), Barker et al. (\protect\hyperlink{ref-barker2015new}{2015}), Burns et al. (\protect\hyperlink{ref-burns2014phylogenetics}{2014}), Hedges et al. (\protect\hyperlink{ref-Hedges2015}{2015}), Hooper and Price (\protect\hyperlink{ref-hooper2017chromosomal}{2017}), Jetz et al. (\protect\hyperlink{ref-Jetz2012}{2012}).) This revealed 28 age data points for the queried species names.}
\label{fig:figure2-1}
\end{figure}
% \begin{center}
% \textsc{Figure \ref{fig:figure2-1}}
% \end{center}
\newpage
\begin{figure}[!h]
\includegraphics{../figures/figure2/figure2-2.pdf}
\caption{}
\label{fig:summaries}
\end{figure}
% \begin{center}
% \textsc{Figure \ref{fig:figure2-2}}
% \end{center}
LTT plots of median (top) and Supermatrix Distance Method (SDM; bottom) chronograms summarising information from source chronograms found for the Fringillidae. Arrows indicate tree maximum age.
\newpage
\begin{figure}[!h]
\includegraphics{../figures/fig_crossval_bladj.pdf}
\caption{}
\label{fig:cvbladj}
\end{figure}
% \begin{center}
% \textsc{Figure \ref{fig:cvbladj}}
% \end{center}
LTT plots showing results from the cross-validation analyses of trees without branch lengths dated using BLADJ. The dating analysis can only be performed in trees with more than 2 tips, thus excluding chronogram from study 4; its data was still used as calibration for the other source chronograms.
\newpage
\begin{figure}[!h]
\includegraphics{../figures/fig_crossval_boldsumm.pdf}
\caption{}
\label{fig:cvbold}
\end{figure}
% \begin{center}
% \textsc{Figure \ref{fig:cvbold}}
% \end{center}
LTT plots showing results from the cross-validation analyses of trees with branch length reconstructed with data from the Barcode of Life Database (BOLD) dated using PATHd8. We could construct a tree with branch lengths for all source chronograms. However, dating with PATHd8 was only successful in three source chronograms shown here.

### 2022-03-24 manuscript/figures.tex
updating figures on ms

% in this file I leave the figure captions outside\ caption{} because I want them
% to be formatted in the same way as the general text (double spaced and linenumbered)
\captionsetup[figure]{labelfont={sc},labelformat={default},labelsep=period,name={Figure}}

\begin{figure}[!h]
\includegraphics{../figures/figure1/figure1-vertical-final.pdf}
\caption{Stylized DateLife algorithm showing the main steps that are performed with the R package \texttt{datelife}, or online at \url{www.datelife.org/query/}. Details on the functions involved on each workflow are available at \texttt{datelife}'s R package manual https://phylotastic.org/datelife/articles/.}
\label{fig:workflow}
\end{figure}
% \begin{center}
% \textsc{Figure \ref{fig:workflow}}
% \end{center}
% Stylized DateLife workflow. This shows the general workflows and analyses that can be performed with \texttt{datelife}, via the R package or through the website  at \url{http://www.datelife.org/}. Details on the functions involved on each workflow are shown in \texttt{datelife}'s R package vignette.
\newpage
\begin{figure}[!h]
\includegraphics[width=1\linewidth]{../figures/fig_runtime_main.pdf}
\caption{Input taxon name processing and chronogram database search computation time increases with number of input taxon names. We sampled N bird species names for each input size class, 100 times, and then performed a \texttt{datelife} search using the Taxon Names Resoultion Service (TNRS; light gray), and without using TNRS (dark gray). We also performed a search using the already processed query for comparison (light blue).}
\label{fig:runtime1}
\end{figure}
% \begin{center}
% \textsc{Figure \ref{fig:runtime1}}
% \end{center}
%
\newpage
\begin{figure}[!h]
\includegraphics[width=1\linewidth]{../figures/fig_schronograms1.pdf}
\caption{Lineage through time (LTT) plots of source chronograms containing all or a subset of species from the bird family Fringillidae of true finches. Arrows indicate maximum age of each chronogram. Numbers reference to chronograms' original publications 1: Barker et al. (\protect\hyperlink{ref-barker2012going}{2012}), 2: Barker et al. (\protect\hyperlink{ref-barker2015new}{2015}), 3: Burns et al. (\protect\hyperlink{ref-burns2014phylogenetics}{2014}), 4: Claramunt and Cracraft (\protect\hyperlink{ref-claramunt2015new}{2015}), 5: Gibb et al. (\protect\hyperlink{ref-gibb2015new}{2015}), 6: Hedges et al. (\protect\hyperlink{ref-Hedges2015}{2015}), 7: Hooper and Price (\protect\hyperlink{ref-hooper2017chromosomal}{2017}), 8: Jetz et al. (\protect\hyperlink{ref-Jetz2012}{2012}), 9: Price et al. (\protect\hyperlink{ref-price2014niche}{2014}).
}
\label{fig:schronograms1}
\end{figure}
% \begin{center}
% \textsc{Figure \ref{fig:schronograms1}}
% \end{center}
%Lineage through time (LTT) plots of source chronograms containing all or a subset of species from the bird family Fringillidae of true finches. Arrows indicate maximum age of each chronogram. Numbers reference to chronograms' original publications 1: Barker et al. (\protect\hyperlink{ref-barker2012going}{2012}), 2: Barker et al. (\protect\hyperlink{ref-barker2015new}{2015}), 3: Burns et al. (\protect\hyperlink{ref-burns2014phylogenetics}{2014}), 4: Claramunt and Cracraft (\protect\hyperlink{ref-claramunt2015new}{2015}), 5: Gibb et al. (\protect\hyperlink{ref-gibb2015new}{2015}), 6: Hedges et al. (\protect\hyperlink{ref-Hedges2015}{2015}), 7: Hooper and Price (\protect\hyperlink{ref-hooper2017chromosomal}{2017}), 8: Jetz et al. (\protect\hyperlink{ref-Jetz2012}{2012}), 9: Price et al. (\protect\hyperlink{ref-price2014niche}{2014}).
\newpage
\begin{figure}[!h]
\includegraphics{../figures/fig_summaries.pdf}
\caption{LTT plots of median (top) and Supermatrix Distance Method (SDM; bottom) chronograms summarising information from source chronograms found for the Fringillidae. Arrows indicate tree maximum age.}
\label{fig:summaries}
\end{figure}
% \begin{center}
% \textsc{Figure \ref{fig:summaries}}
% \end{center}
% LTT plots of median (top) and Supermatrix Distance Method (SDM; bottom) chronograms summarising information from source chronograms found for the Fringillidae. Arrows indicate tree maximum age.
\newpage
\begin{figure}[!h]
\includegraphics{../figures/fig_crossval_bladj.pdf}
\caption{LTT plots showing results from the cross-validation analyses of trees without branch lengths dated using BLADJ. The dating analysis can only be performed in trees with more than 2 tips, thus excluding chronogram from study 4; its data was still used as calibration for the other source chronograms.}
\label{fig:cvbladj}
\end{figure}
% \begin{center}
% \textsc{Figure \ref{fig:cvbladj}}
% \end{center}
% LTT plots showing results from the cross-validation analyses of trees without branch lengths dated using BLADJ. The dating analysis can only be performed in trees with more than 2 tips, thus excluding chronogram from study 4; its data was still used as calibration for the other source chronograms.
\newpage
\begin{figure}[!h]
\includegraphics{../figures/fig_crossval_boldsumm.pdf}
\caption{LTT plots showing results from the cross-validation analyses of trees with branch length reconstructed with data from the Barcode of Life Database (BOLD) dated using PATHd8. We could construct a tree with branch lengths for all source chronograms. However, dating with PATHd8 was only successful in three source chronograms shown here.}
\label{fig:cvbold}
\end{figure}
% \begin{center}
% \textsc{Figure \ref{fig:cvbold}}
% \end{center}
FIGURES
%LTT plots showing results from the cross-validation analyses of trees with branch length reconstructed with data from the Barcode of Life Database (BOLD) dated using PATHd8. We could construct a tree with branch lengths for all source chronograms. However, dating with PATHd8 was only successful in three source chronograms shown here.

%\end{linenumbers}

>>>>

\captionsetup[figure]{labelfont={sc},labelformat={default},labelsep=period,name={Figure}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% WORKFLOW FIGURE
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{figure}[!h]
\includegraphics{../figures/figure1/figure1-vertical-final.pdf}
\caption{
Stylized DateLife algorithm showing the main steps that are performed
with the R package \texttt{datelife}, or online at \url{www.datelife.org/query/}.
Details on the functions involved on each workflow are available at \texttt{datelife}'s
R package manual https://phylotastic.org/datelife/articles/.
}
\label{fig:workflow}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% BENCHMARK FIGURE
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
\begin{figure}[!h]
\includegraphics{../figures/fig_runtime_main.pdf}
\caption{Input taxon name processing and chronogram database search computation time increases with number of input taxon names. We sampled N bird species names for each input size class, 100 times, and then performed a \texttt{datelife} search using the Taxon Names Resoultion Service (TNRS; light gray), and without using TNRS (dark gray). We also performed a search using the already processed query for comparison (light blue).}
\label{fig:benchmark}
\end{figure}
\newpage

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% SMALL EXAMPLE FIGURES
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newpage
\begin{figure}[!h]
\includegraphics[width=1\linewidth]{../figures/figure2/figure2-1.pdf}
\caption{
Age data results of a DateLife search of a small sample of 6 bird species within the Passeriformes. Input names were found across 9 chronograms within 6 independent studies (Barker et al. (\protect\hyperlink{ref-barker2012going}{2012}), Barker et al. (\protect\hyperlink{ref-barker2015new}{2015}), Burns et al. (\protect\hyperlink{ref-burns2014phylogenetics}{2014}), Hedges et al. (\protect\hyperlink{ref-Hedges2015}{2015}), Hooper and Price (\protect\hyperlink{ref-hooper2017chromosomal}{2017}), Jetz et al. (\protect\hyperlink{ref-Jetz2012}{2012}).) This revealed 28 age data points for the queried species names.
}
\label{fig:figure2-1}
\end{figure}

\newpage
\begin{figure}[!h]
\includegraphics[width=1\linewidth]{../figures/figure2/figure2-2.pdf}
\caption{
Summarized age data is used as secondary calibrations to obtain a summary chronogram.
}
\label{fig:figure2-2}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% REAL LIFE APP FIGURE
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newpage
\begin{figure}[!h]
\includegraphics{../figures/fig_summaries.pdf}
\caption{LTT plots of median (top) and Supermatrix Distance Method (SDM; bottom) chronograms summarising information from source chronograms found for the Fringillidae. Arrows indicate tree maximum age.}
\label{fig:summaries}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% CROSS VALIDATION FIGURES
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newpage
\begin{figure}[!h]
\includegraphics{../figures/fig_crossval_bladj.pdf}
\caption{LTT plots showing results from the cross-validation analyses of trees without branch lengths dated using BLADJ. The dating analysis can only be performed in trees with more than 2 tips, thus excluding chronogram from study 4; its data was still used as calibration for the other source chronograms.}
\label{fig:cvbladj}
\end{figure}

\newpage
\begin{figure}[!h]
\includegraphics{../figures/fig_crossval_boldsumm.pdf}
\caption{LTT plots showing results from the cross-validation analyses of trees with branch length reconstructed with data from the Barcode of Life Database (BOLD) dated using PATHd8. We could construct a tree with branch lengths for all source chronograms. However, dating with PATHd8 was only successful in three source chronograms shown here.}
\label{fig:cvbold}
\end{figure}

### 2022-03-24 manuscript/manuscript.Rmd
updating figures on ms

documentclass     : "apa6"
classoption       : "man"
output            :
  papaja::apa6_pdf:
    includes      :
      in_header: "preamble.tex"
      after_body  : "figures.tex"

nocite: |
  @barker2012going, @barker2015new, @burns2014phylogenetics, @claramunt2015new, @gibb2015new, @Hedges2015, @hooper2017chromosomal, @Jetz2012, @price2014niche
---

@@

# Introduction
<!-- If you're aiming for syst bio you don't want to start so general Readers know what chronorgrams and phylogenies are!.-->
Chronograms --phylogenies with branch lengths proportional to time-- provide key data for the study of natural processes in many areas of biological research, such as developmental  biology [@laubichler2009form; @delsuc2018phylogenomic], conservation biology [@Felsenstein1985a; @Webb2000], historical biogeography [@posadas2006historical], and species diversification [@magallon2001absolute; @Morlon2014].

@@

DateLife can also reconstruct branch lengths proportional to substitution rates on the fixed tree topology using available genetic data from BOLD.

@@

_**Summarizing node ages.--**_ DateLife summarizes matched calibrations  ??? How is 'calibrations' defined here? the calibrations from congruification, or the pairwise distances??? into a single patristic distance matrix using different methods. 
Summarizing options implemented include Super Distance Matrix method [SDM, @Criscuolo2006] and summary statistics such as median, minimum and maximum ages.
For each node in the output tree topology, the pairwise distances that traverse that node are ..???
Under the SDM criterion ??? what happens??
Using the minimum age criterion ..??

@@

`datelife`'s code speed was tested on an Apple iMac
with one 3.4 GHz Intel Core i5 processor. <!--No multiprocessor tests were performed-->
We registered variation in computing time of query processing and search through the database relative to number of queried taxon names.
Query processing time increases roughly linearly with number of input taxon names, and
increases considerably if TNRS is activated.
<!-- Defining tnrs here again, bc we only talk about it two times, so it easy to forget it -->
Up to ten thousand names can be processed and searched in less than 30 minutes with the most time consuming settings.
Once names have been processed as described in methods, a name search through the chronogram database can be performed in less than a minute, even with a very large number of taxon names (Fig. \ref{fig:runtime1}).
<!--Speed of other datelife workflows? NO, bc it reflects speed from other software mainly. But I will put a sentence about it...
Summarizing `datelife` results processing times: shows sdm speed, clustering methods speed;
generating new trees: shows bladj speed;
Adding dates processing time: reflects more mrbayes speed than anything;
get_bold_otol_tree running time: reflects bold downloading data speed.-->
<!--Speed with different number of chronograms in database? NO-->
<!--Speed of web interface and of r package in computers with different capacities? NOOOO-->
`datelife`'s code performance was evaluated with a set of unit tests designed and
implemented with the R package testthat [@RCoreTeam2018] that were run both locally
with the devtools package [@RCoreTeam2018], and on a public server --via
GitHub, using the continuous integration tool Travis CI (<https://travis-ci.org>). At
present, unit tests cover more than 30% of `datelife`'s code (<https://codecov.io/gh/phylotastic/datelife>). <!-- unit test make code robust, verifiable, debuggable, -->

@@

We illustrate the `datelife` algorithm using the family of true finches, Fringillidae as an example.
<!--We used `r cite_r("r-references.bib")` for all our analyses.-->
A college educator wishes to obtain state-of-the-art data on time of evolutionary origin of species belonging to the true finches for their class. They decide to use `datelife` because they are teaching best practices for reproducibility.
Students have the option to go to the website at www.datelife.org and perform an interactive run. However, the educator also wants the students to practice their R skills.
The first step is to run a `datelife` query using the "get species from taxon" flag. This will get all recognised species names within their chosen inclusive taxon. The Fringillidae has `r length(dquery$cleaned_names)` species, according to the Open Tree of Life taxonomy.
Once with a curated set of species taxon names, the next step is to run a `datelife` search that will find all chronograms that contain at least two species names. The algorithm proceeds to prune the trees to keep matching species names on tips only, and transform the pruned trees to pairwise distance matrices.
 There are `r length(dres)` chronograms containing at least two Fringillidae species, published in `r length(unique(names(dres)))` different studies (Fig. \ref{fig:schronograms1}).
<!-- In case we want to cite the studies in here, but they are already cited in the figure caption: [Fig. \ref{fig:schronograms1}; @barker2012going; @barker2015new; @burns2014phylogenetics; @claramunt2015new; @gibb2015new; @Hedges2015; @hooper2017chromosomal; @Jetz2012; @price2014niche]. -->
The final step is to summarize the available information using two alternative types of summary chronograms, median and SDM. As explained in the "Description" section, data from source chronograms is first summarised into a single distance matrix and then the available node ages are used as fixed node calibrations over a consensus tree topology, to obtain a fully dated tree with the program BLADJ (Fig. \ref{fig:summaries}). Median summary chronograms are older and have wider variation in maximum ages than chronograms obtained with SDM. 
?? Say some things about the results! 
<!-- (Going from a summary distance matrix Using this method is better than using clustering algorithms  -->
<!-- and hierarchical clustering method which appears to push ages back -->

>>>>

documentclass     : "apa6"
classoption       : "man"
output            :
  papaja::apa6_pdf:
    includes      :
      in_header: "preamble.tex"
      after_body  : "figures-alt.tex"

nocite: |
  @barker2013going, @barker2015new, @burns2014phylogenetics, @claramunt2015new, @gibb2015new, @Hedges2015, @hooper2017chromosomal, @Jetz2012, @price2014niche
---

@@

# Introduction
<!-- If you're aiming for syst bio you don't want to start so general Readers know what chronograms and phylogenies are!.-->
Chronograms --phylogenies with branch lengths proportional to time-- provide key data for the study of natural processes in many areas of biological research, such as developmental biology [@laubichler2009form; @delsuc2018phylogenomic], conservation biology [@Felsenstein1985a; @Webb2000], historical biogeography [@posadas2006historical], and species diversification [@magallon2001absolute; @Morlon2014].

@@

DateLife can also reconstruct branch lengths proportional to substitution rates on a fixed tree topology using available genetic data from BOLD.

@@

_**Summarizing node ages.--**_ DateLife summarizes matched calibrations
<!--??? How is 'calibrations' defined here? the calibrations from congruification, or the pairwise distances??? -->
into a single patristic distance matrix using different methods.
Summarizing options implemented include Super Distance Matrix method [SDM, @Criscuolo2006] and summary statistics such as median, minimum and maximum ages.
<!--
For each node in the output tree topology, the pairwise distances that traverse that node are ..???
Under the SDM criterion ??? what happens??
Using the minimum age criterion ..??
-->

@@

`datelife`'s code speed was tested on an Apple iMac
with one 3.4 GHz Intel Core i5 processor. <!--No multiprocessor tests were performed-->
We registered variation in computing time of query processing and search through the database relative to number of queried taxon names.
Query processing time increases roughly linearly with number of input taxon names, and
increases considerably if TNRS is activated.
<!-- Defining tnrs here again, bc we only talk about it two times, so it easy to forget it -->
Up to ten thousand names can be processed and searched in less than 30 minutes with the most time consuming settings.
Once names have been processed as described in methods, a name search through the chronogram database can be performed in less than a minute, even with a very large number of taxon names (Fig. \ref{fig:benchmark}).
<!--Speed of other datelife workflows? NO, bc it reflects speed from other software mainly. But I will put a sentence about it...
Summarizing `datelife` results processing times: shows sdm speed, clustering methods speed;
generating new trees: shows bladj speed;
Adding dates processing time: reflects more mrbayes speed than anything;
get_bold_otol_tree running time: reflects bold downloading data speed.-->
<!--Speed with different number of chronograms in database? NO-->
<!--Speed of web interface and of r package in computers with different capacities? NOOOO-->
`datelife`'s code performance was evaluated with a set of unit tests designed and
implemented with the R package testthat [@RCoreTeam2018] that were run both locally
with the devtools package [@RCoreTeam2018], and on a public server --via
GitHub, using the continuous integration tool Travis CI (<https://travis-ci.org>). At
present, unit tests cover more than 40% of `datelife`'s code (<https://codecov.io/gh/phylotastic/datelife>). <!-- unit test make code robust, verifiable, debuggable, -->

@@

We illustrate the DateLife algorithm using a group within the Passeriform birds encompassing the family of true finches, Fringillidae and allies as case study. The first example analyses 6 bird species and shows all steps of the algorithm. The second example is a real life application
<!--We used `r cite_r("r-references.bib")` for all our analyses.-->

## Small example

We chose 6 bird species associated to true finches at random. The sample includes
two species of cardinals:
the black-thighed grosbeak -- *Pheucticus tibialis* and
the crimson-collared grosbeak -- *Rhodothraupis celaeno*;
three species of buntings:
the yellowhammer -- *Emberiza citrinella*,
the pine bunting --  *Emberiza leucocephalos* and
the yellow-throated bunting --  *Emberiza elegans*;
and one species of tanager, the vegetarian finch -- *Platyspiza crassirostris*.

Processing input names found that *Emberiza elegans* is synonym for *Schoeniclus elegans* in the reference taxonomy.
DateLife used the processed input names to search the local chronogram database and found 9 matching chronograms in 6 different studies. Three studies matched five input names [@barker2015new; @Hedges2015; @Jetz2012], one study matched four input names [@hooper2017chromosomal] and two studies matched two input names [@barker2013going; @burns2014phylogenetics]. No studies matched all input names. Together, matching chronograms have 28 unique age data points. All nodes have age data.
As fixed tree topology, DateLife used OpenTree's synthetic tree as default and mapped age data to nodes in the tree.
As expected, more inclusive nodes (e.g., node "n1") have more age data than less inclusive nodes (e.g., node "n5").
The processing step allowd discovering five data points for node "n4" that would not have had any data otherwise.
Age summary statistics per node were calculated and tested as secondary calibrations to date the tree topology using the BLADJ algorithm.
Age data for node "n2" was excluded as final calibration because it is older than age data of a more inclusive node.

## Real life application

A college educator wishes to obtain state-of-the-art data on time of evolutionary origin of species belonging to the true finches for their class. They decide to use `datelife` because they are teaching best practices for reproducibility.
Students have the option to go to the website at www.datelife.org and perform an interactive run. However, the educator also wants the students to practice their R skills.
The first step is to run a `datelife` query using the "get species from taxon" flag. This will get all recognised species names within their chosen inclusive taxon. The Fringillidae has `r length(dquery$cleaned_names)` species, according to the Open Tree of Life taxonomy.
Once with a curated set of species taxon names, the next step is to run a `datelife` search that will find all chronograms that contain at least two species names. The algorithm proceeds to prune the trees to keep matching species names on tips only, and transform the pruned trees to pairwise distance matrices.
 There are `r length(dres)` chronograms containing at least two Fringillidae species, published in `r length(unique(names(dres)))` different studies (Fig. \ref{fig:schronograms1}).
<!-- In case we want to cite the studies in here, but they are already cited in the figure caption: [Fig. \ref{fig:schronograms1}; @barker2013going; @barker2015new; @burns2014phylogenetics; @claramunt2015new; @gibb2015new; @Hedges2015; @hooper2017chromosomal; @Jetz2012; @price2014niche]. -->
The final step is to summarize the available information using two alternative types of summary chronograms, median and SDM. As explained in the "Description" section, data from source chronograms is first summarised into a single distance matrix and then the available node ages are used as fixed node calibrations over a consensus tree topology, to obtain a fully dated tree with the program BLADJ (Fig. \ref{fig:summaries}). Median summary chronograms are older and have wider variation in maximum ages than chronograms obtained with SDM.
?? Say some things about the results!
<!-- (Going from a summary distance matrix Using this method is better than using clustering algorithms  -->
<!-- and hierarchical clustering method which appears to push ages back -->
